### 2018-03-17 paper/5-implementations.tex
More work on S5

We provide the complete implementation of \Make in our framework in
Fig.~\ref{fig-make-implementation}. \Make processes keys in a linear order based
on the topological sort (see~\S\ref{sec-concurrency} for parallel \Make). For
each key, it builds it if it is older than any of its dependencies. We capture
the persistent build information that \Make stores by a pair
\hs{(@@modTime,}~\hs{now)} comprising the \emph{file modification time} function
\hs{modTime}~\hs{::}~\hs{k}~\hs{->}~\hs{Time} and the \emph{current time} \hs{now}.
Setting aside the explicit manipulation of file modification times, which in
reality is taken care of by the file system, the function \hs{make} captures
the essence of \Make in a clear and precise manner.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Persistent build information
type Time       = Integer
type MakeInfo k = (k -> Time, Time)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Eq k => Build Applicative (MakeInfo k) k v
make = topological process
  where
    process key deps act = do
        (modTime, now) <- getInfo <$> get
        let dirty = or [ modTime dep > modTime key | dep <- deps ]
        when dirty $ do
            v <- act
            let newModTime k = if k == key then now else modTime k
            modify $ \s -> putInfo (putValue s key v) (newModTime, now + 1)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
transitiveClosure :: Eq a => (a -> [a]) ->  a  -> [a]
topologicalSort   :: Eq a => (a -> [a]) -> [a] -> [a]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Topological dependency strategy
topological :: Eq k => (k -> [k] -> State (Store i k v) v -> State (Store i k v) (@@))
            -> Build Applicative i k v
topological process task key = execState $ forM_ chain $ \k -> do
    let fetch k = do store <- get; return (getValue store k)
    case task fetch k of
        Nothing  -> return ()
        Just act -> process k (deps k) act
  where
    deps  = dependencies task
    chain = topologicalSort deps (transitiveClosure deps key)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-2mm}
\end{figure}

The helper function \hs{topological} calls \hs{process} on every \hs{key} in
the topological order, providing the list of its dependencies \hs{deps} and the
action \hs{act} to compute the resulting value if it needs to be rebuilt. To
determine if the \hs{key} is \hs{dirty}, \hs{process} compares its modification
time with those of its dependencies. If the \hs{key} needs to be rebuilt, the
action \hs{act} is executed and the obtained result is stored, along with an
updated file modification timestamp.

The implementation of \hs{topological} encodes the dependency strategy that
\Make has chosen to use. The \hs{where} clause corresponds to the pre-processing
stage, which uses the function \hs{dependencies}, defined in~\S\ref{sec-deps},
to extract static dependencies from the provided applicative \hs{task}. We
compute the linear processing \hs{chain} by taking the \hs{transitiveClosure}
of \hs{key}'s dependencies, and performing the \hs{topologicalSort} of the
result. We omit implementation of textbook graph algorithms
\hs{transitiveClosure} and \hs{topologicalSort}, e.g.
see~\cite{cormen2001introduction}.

@@

\todo{AM}{Note that anything updating input files must also update Make's info.}
\todo{AM}{Say that in reality that build rules mutate the state themselves.}

@@

As a programming exercise, \Excel is the most challenging, due to the
complicated reordering dependency strategy that requires retrying out-of-order
operations, see Fig.~\ref{fig-excel-implementation}.

\Excel's persistently stored information is a triple: (i) the dirty bit
function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
\hs{k}~\hs{->}~\hs{DependencyApproximation}~\hs{k} that \Excel uses to handle
dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
previous build (\S\ref{sec-background-excel}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Approximation of task dependencies
data DependencyApproximation k = SubsetOf [k] | Unknown
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Persistent build information
type CalcChain k = [k]
type ExcelInfo k = ((k -> Bool, k -> DependencyApproximation k), CalcChain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Result of speculative task execution
data Result k v = MissingDependency k | Result v [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Reordering dependency strategy (implementation omitted)
reordering :: Ord k
    => (k -> State (Store i k v) (Result k v) -> State (Store i k v) (Maybe (Result k v)))
    -> Build Monad (i, CalcChain k) k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = reordering process
  where
    process key act = do
        (dirty, deps) <- getInfo <$> get
        let rebuild = dirty key || case deps key of SubsetOf ks -> any dirty ks
                                                    Unknown     -> True
        if not rebuild
            then return Nothing
            else do
                result <- act
                case result of
                    MissingDependency _ -> return ()
                    Result v _dynamicDependencies -> do
                        let newDirty k = if k == key then True else dirty k
                        modify $ \s -> putInfo (putValue s key v) (newDirty, deps)
                return (Just result)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2mm}
\end{figure}

The helper function \hs{reordering}, whose 21-line implementation we omit since it is
technical and not particularly enlightening, calls the function \hs{process} to
\emph{try to build} a \hs{key} by executing the action \hs{act}, in the order
determined by the calculation chain. To decide whether to \hs{rebuild} the
\hs{key}, \hs{process} checks if the \hs{key} itself is marked dirty or the
approximation of its dependencies contains a dirty key. Notice that if the
dependencies of the \hs{key} are \hs{Unknown}, e.g. when it uses the
\cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
not needed, \hs{process} returns \hs{Nothing} to indicate this. Otherwise, it
executes \hs{act} leading to two possible \hs{result}s:

@@

In both of the above cases, we notify the parent \hs{reordering} function about
the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
that \hs{process} ignores \hs{dynamicDependencies} available in the \hs{result}.
While not required for \Excel, we have implemented build systems using
\hs{reordering} which use verifying and constructive traces, effectively turning
\Excel into a cloud build system and ensuring \hs{reordering} is not overly
fitted to \Excel alone (additional implementations will be submitted to the
Artefact Evaluation).

@@

\begin{description}
\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the \hs{key} you are currently building, and a function \hs{check} which checks a specified dependency. It returns the \hs{result} field of all traces that match. Since \hs{check} is in an arbitrary monad, the function has to use \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&} functions\footnote{These functions are all available in the \cmd{extra} library on Hackage.}.
\item[\hs{recursive}] defines the dependency ordering pattern. It requires a \hs{process} function that builds a key given a way to build a dependency, and a way to build the result. The main purpose of \hs{recursive} is to ensure that in a single run any key is not built twice -- for which it extends the \hs{State} monad with a list of \hs{done} keys.
\item[\hs{shake}] ties everything together. First it tests if the traces allow the current state of the target key. If not, it builds the key and updates the store. The only subtlty is that Shake calls \hs{fetch} on the keys while checking them -- building the last-recorded dependencies before checking them. One minor annoyance is that the \hs{State} has been extended and thus needs to be projected using \hs{fst} before it is used.
\end{description}

>>>>

We provide an implementation of \Make using our framework in
Fig.~\ref{fig-make-implementation}. \Make processes keys in a linear order based
on a topological sort (see~\S\ref{sec-concurrency} for parallel \Make). For
each key, it builds it if it is older than any of its dependencies. We capture
the persistent build information that \Make stores by a pair
\hs{(@@modTime,}~\hs{now)} comprising the \emph{file modification time} function
\hs{modTime}~\hs{::}~\hs{k}~\hs{->}~\hs{Time} and the \emph{current time} \hs{now}.
Setting aside the explicit manipulation of file modification times, which in
reality is taken care of by the file system, the function \hs{make} captures
the essence of \Make in a clear and precise manner.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Persistent build information
type Time       = Integer
type MakeInfo k = (k -> Time, Time)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Eq k => Build Applicative (MakeInfo k) k v
make = topological process
  where
    process key deps act = do
        (modTime, now) <- gets getInfo
        let dirty = or [ modTime dep > modTime key | dep <- deps ]
        when dirty $ do
            v <- act
            let newModTime k = if k == key then now else modTime k
            modify $ putInfo (newModTime, now + 1) . putValue key v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
transitiveClosure :: Eq a => (a -> [a]) ->  a  -> [a]
topologicalSort   :: Eq a => (a -> [a]) -> [a] -> [a]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Topological dependency strategy
topological :: Eq k => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
            -> Build Applicative i k v
topological process task key = execState $ forM_ chain $ \k -> do
    let fetch k = gets (getValue k)
    case task fetch k of
        Nothing  -> return ()
        Just act -> process k (deps k) act
  where
    deps  = dependencies task -- dependencies is defined in @\S\ref{secdeps}@
    chain = topSort deps (reachable deps key)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-2mm}
\end{figure}

The helper function \hs{topological} calls \hs{process} on every \hs{key} in
a topological order, providing the list of its dependencies \hs{deps} and the
action \hs{act} to compute the resulting value if it needs to be rebuilt. To
determine if the \hs{key} is \hs{dirty}, \hs{process} compares its modification
time with those of its dependencies. If the \hs{key} needs to be rebuilt, the
action \hs{act} is executed and the obtained result is stored, along with an
updated file modification timestamp.

The implementation of \hs{topological} encodes the dependency strategy that
\Make has chosen to use. The \hs{where} clause corresponds to the pre-processing
stage, which uses the function \hs{dependencies}, defined in~\S\ref{sec-deps},
to extract static dependencies from the provided applicative \hs{task}. We
compute the linear processing \hs{chain} by taking the keys \hs{reachable}
from \hs{key} via dependencies, and performing the topological sort of the
result. We omit implementation of textbook graph algorithms
\hs{reachable} and \hs{topSort}, e.g.
see~\cite{cormen2001introduction}.

@@

@@

We define \Excel, with it's reordering dependency strategy, in  Fig.~\ref{fig-excel-implementation}.
\Excel's persistently stored information is a triple: (i) the dirty bit
function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
\hs{k}~\hs{->}~\hs{DependencyApproximation}~\hs{k} that \Excel uses to handle
dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
previous build (\S\ref{sec-background-excel}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Approximation of task dependencies
data DependencyApproximation k = SubsetOf [k] | Unknown
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Persistent build information
type CalcChain k = [k]
type ExcelInfo k = ((k -> Bool, k -> DependencyApproximation k), CalcChain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Result of speculative task execution
data Result k v = MissingDependency k | Result v [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Reordering dependency strategy (implementation omitted, 21 lines)
reordering :: Ord k
    => (k -> State (Store i k v) (Result k v) -> State (Store i k v) (Maybe (Result k v)))
    -> Build Monad (i, CalcChain k) k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = reordering process
  where
    process key act = do
        (dirty, deps) <- gets getInfo
        let rebuild = dirty key || case deps key of SubsetOf ks -> any dirty ks
                                                    Unknown     -> True
        if not rebuild
            then return Nothing
            else do
                result <- act
                case result of
                    MissingDependency _ -> return ()
                    Result v _dynamicDependencies -> do
                        let newDirty k = if k == key then True else dirty k
                        modify $ putInfo (newDirty, deps) . putValue key v
                return (Just result)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2mm}
\end{figure}

The helper function \hs{reordering}, whose implementation we omit since it is
technical and not particularly enlightening, calls the function \hs{process} to
\emph{try to build} a \hs{key} by executing the action \hs{act}, in the order
determined by the calculation chain. To decide whether to \hs{rebuild} the
\hs{key}, \hs{process} checks if the \hs{key} itself is marked dirty or the
approximation of its dependencies contains a dirty key. Notice that if the
dependencies of the \hs{key} are \hs{Unknown}, e.g. when it uses the
\cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
not needed, \hs{process} returns \hs{Nothing} to indicate that. Otherwise, it
executes \hs{act} leading to one of two possible \hs{result}s:

@@

In both of the above cases, we notify the parent \hs{reordering} function of
the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
that \hs{process} ignores \hs{dynamicDependencies} available in the \hs{result}.
While not required for \Excel, we have implemented build systems using
\hs{reordering} which use verifying and constructive traces, effectively turning
\Excel into a cloud build system and ensuring \hs{reordering} is not overly
fitted to \Excel alone (additional implementations will be submitted to the
Artefact Evaluation).

In reality \Excel works slightly differently (as far as we are able to ascertain) --
on a change it propagates the dirty bits forward using the \hs{dynamicDependencies}, then only checks if the current \hs{key} is dirty. While both methods are equivalent, merely changing the interleaving, our approach allows us to model more of the behaviour of \Excel.

@@

\begin{description}
\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the \hs{key} you are currently building, and a function \hs{check} which checks a specified dependency. It returns the \hs{result} field of all traces that match. Since \hs{check} is in an arbitrary monad, the function has to use \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&} functions\footnote{These functions are all available in the \cmd{extra} library on Hackage.}.
\item[\hs{recursive}] defines the dependency ordering pattern. It requires a \hs{process} function that builds a key given a way to recursively build a dependency, and a way to run task to produce result. The main purpose of \hs{recursive} is to ensure that in a single run any key is not built twice -- for which it extends the \hs{State} monad with a list of \hs{done} keys.
\item[\hs{shake}] ties everything together. First it tests if the traces allow the current state of the target key. If not, it builds the key and updates the store. The only subtlty is that Shake calls \hs{fetch} on the keys while checking them -- building the last-recorded dependencies before checking them. One minor annoyance is that the \hs{State} has been extended and thus needs to be projected using \hs{fst} before it is used.
\end{description}
